---
title: "Diabetes"
output: html_document
date: "2025-10-22"
---

```{r include=FALSE}
install.packages("dplyr")
library(dplyr)
```

```{r echo=FALSE}
#Loading the dataset
diabetes = read.csv("diabetes.csv", head=TRUE, sep=",")
summary(diabetes)
```

# Step 1 : Data Cleaning and Preparation

First we are going to replace all the impossible values, such as :

-   the Glucose level at 0

-   the Blood Pressure at 0

-   the Skin Thickness at 0

-   the BMI at 0

-   and the Diabetes Pedigree Function at 0
We obtain the following summary :

```{r echp=FALSE}
#mutating each columns concern, if a value equals 0, it's replaced by an NA value
diabetes <- diabetes %>%
  mutate(
    Glucose = na_if(Glucose, 0),
    BloodPressure = na_if(BloodPressure, 0),
    SkinThickness = na_if(SkinThickness, 0),
    BMI = na_if(BMI, 0),
    DiabetesPedigreeFunction = na_if(DiabetesPedigreeFunction, 0),
  )
summary(diabetes)
```
Now we are going to replace those NA values by the median of each columns.
We obtain the following summary :

<<<<<<< HEAD
```{r echo=FALSE}
#mutating each columns concern, if a value equals NA, it's replaced by the median of the corresponding columns else it stays the same
=======
```{r}
>>>>>>> a5820b9a41e14688a5455e2bd83477a1fc6f7871
diabetes <- diabetes %>%
  mutate(
    Glucose = if_else(is.na(Glucose), median(Glucose, na.rm = TRUE), Glucose),
    BloodPressure = if_else(is.na(BloodPressure), median(BloodPressure, na.rm = TRUE), BloodPressure),
    SkinThickness = if_else(is.na(SkinThickness), median(SkinThickness, na.rm = TRUE), SkinThickness),
    BMI = if_else(is.na(BMI), median(BMI, na.rm = TRUE), BMI),
    DiabetesPedigreeFunction = if_else(is.na(DiabetesPedigreeFunction), median(DiabetesPedigreeFunction, na.rm = TRUE), DiabetesPedigreeFunction)
  )

summary(diabetes)
```
<<<<<<< HEAD
Unlike the previous summary we can see in this one that we don't have any NA Values left
=======

```{r}
diabetes_var<-colnames(diabetes)
diabetes_var_without_outcome<-diabetes_var[diabetes_var!="Outcome"]
diabetes_scaled <-diabetes %>% mutate_at(diabetes_var_without_outcome, ~(scale(.) %>% as.vector))
diabetes_scaled
```
>>>>>>> a5820b9a41e14688a5455e2bd83477a1fc6f7871
